---
title: "AEMO Analysis"
author: "Oliver Sheridan-Methven"
date: "25^th^ September 2018"
header-includes:
   - \usepackage{bm}
   - \usepackage{bbm}
   - \usepackage{physics}
   - \usepackage{cleveref}
   - \usepackage{natbib}
fontsize: 12pt
output: 
    pdf_document: 
        fig_caption: yes
---


# Abstract 

In this report we investigate the relationship between electricity prices
and daytime temperatures in Sydney Australia during 2013. We build a
second order non-linear model based on the minimum daytime temperature,
and predict the prices for the first week in January. This model is able to 
capture weekend and weekday differences.

# The Raw Data

```{r, echo=FALSE, include=FALSE}
library(knitr)  # For making a nice markdown table. 
library(plyr)
# Reading in the data:
AEMO <- read.csv("AEMOPriceDataSimple.csv")
SydTemp <- read.csv("SydTemp.csv")
minTemp <- SydTemp$Minimum.temperature..Degree.C.
maxTemp <- SydTemp$Maximum.temperature..Degree.C.
maxTemp <- maxTemp[SydTemp$Year == 2013]
minTemp <- minTemp[SydTemp$Year == 2013]
maxTemp <- maxTemp[!is.na(maxTemp)]
minTemp <- minTemp[!is.na(minTemp)]

Date <- as.Date(AEMO$Date, '%d/%m/%y')
AEMO <- subset(AEMO, select = c("RRP"))
AEMO <- cbind(AEMO, Date)

SydTemp <-
cbind(SydTemp, Date = as.Date(ISOdate(SydTemp$Year, SydTemp$Month, SydTemp$Day)))

df <- na.omit(merge(AEMO, SydTemp, by.x = "Date", by.y = "Date"))
df <- rename(df, c("Maximum.temperature..Degree.C."="temp_max","Minimum.temperature..Degree.C."="temp_min"))
df <- subset(df, select = c("temp_max", "temp_min", "Date", 'RRP'))
example_data <- cbind(head(df)) # cbind ensures we have a (deep) copy.
```

The data available is the RRP (the electricity price measured in AUD \$), 
and the daily maximum and minimum temperature values, 
where some example data 
is shown in \Cref{tab:tab_example_data}.

```{r tab_example_data, echo=FALSE}
library(data.table) # For renaming some data column names. 
setnames(example_data, old=c("temp_max","temp_min"), new=c("Maximum Temperature", "Minimum Temperature"))
kable(example_data, align='cccc', caption = "\\label{tab:tab_example_data}Some example data illustrating the fields/features available for fitting.")
```

## Outliers

There is a notable outlier in the electricity prices for 20^th^ December 2013, 
(caused by an electrical storm). Due to the esoteric nature of such an event,
we do not want to consider such a point for use in everyday price predictions, 
so we exclude this from our considerations. The resultant dependence between the 
electricity prices and the minimum daily temperature is shown in 
 \Cref{fig:fig_prices_max_temp}. 

```{r, echo=FALSE, include=FALSE}
# we notice this has an obvious outlier in December
df[df$RRP > 200, ]
# we see this corresponds to the large electrical storm on 20th December 2013, so will delete
# this from our dataset.
df <- df[df$RRP < 200, ]
plot(df$Date, df$RRP, xlab = "Date", ylab = "RRP", title("Electricity Prices"))
# We notice that this perhaps looks a little buldge in Australia's Winter months (May-September)
summary(df$RRP)
```

```{r fig_prices_max_temp, echo=FALSE, fig.cap="\\label{fig:fig_prices_max_temp}The dependence on the electricity prices with variations in the daily temperature minimum."}
plot(df$temp_min, df$RRP, xlab = "Minimum Temperature (C)", ylab = "RRP (AUD $)", title("Electricity Prices"))
```

# Non-linear temperature effects 

We recall the linear model
\begin{equation}
\label{eqt:linear_model}
Y = X\beta + \epsilon.
\end{equation}

```{r, echo=FALSE, include=FALSE}
# Pulling out the p_value
linearModel <- lm(df$RRP~df$temp_min)
linear_model_p_value <- anova(linearModel)$'Pr(>F)'[1]
nonlinearModel <- lm(df$RRP~df$temp_min + I(df$temp_min^2))
non_linear_model_p_value <- coef(summary(nonlinearModel))[, "Pr(>|t|)"]
```
To assess whether non-linear temperature effects are significant, we fit a 
linear model and a second order polynomial model to the data. We find the 
linear model \eqref{eqt:linear_model} has the $p$-value $`r signif(linear_model_p_value, 2)`$, 
 which is significant. Furthermore, trying a higher order non-linear model
with quadratic terms achieves $p$-values of $`r signif(non_linear_model_p_value, 2)`$, for the
intercept, linear, and quadratic terms respectively.
We can see these $p$-values are less than 0.05, and so we believe they are 
statistically significant to a 95\% confidence level (approximately a $2\sigma$ 
confidence). 


# Building model matrices

We can build model matrices using the following R-function: 

```{r}
pieLinear <- function(x, k)
{
  # Produces a model matrix for linear models using knots.
  positive <- function(x)
  {
    # Replaces negative values with zeros.
    x[x < 0] <- 0
    return(x)
  }
  
  modelMatrix <- matrix(c(x))
  for (i in 1:length(k))
  {
    modelMatrix <- cbind(modelMatrix, positive(x - k[i]))
  }
  return(modelMatrix)
}
```

Which when given $\bm{x} = (1,2,\ldots,10)$ and knot points $(2,6)$ will output the
model matrix:

```{r}
x <- 1:10
knots <- c(2, 6)
X <- pieLinear(x, knots)
print(X)
```

