---
title: "AEMO Analysis"
author: "Oliver Sheridan-Methven"
date: "15^th^ May 2019"
fontsize: 12pt
header-includes:
   - \usepackage{bm}
   - \usepackage{bbm}
   - \usepackage{physics}
   - \usepackage{cleveref}
   - \usepackage{natbib}
   - \usepackage{textcomp}
   - \usepackage[textsize=small,disable]{todonotes}
   - \usepackage{geometry}
   - \geometry{right=55mm, left=10mm, top=25mm, bottom=25mm, marginparwidth=50mm}
output: 
    pdf_document: 
        fig_caption: yes
---

```{r setup, include=FALSE}
library(reticulate) # Allowing us to use python code in RMarkdown. 
library(knitr) # For including tables
# ADVANCED: using the non default python interpreter. 
use_python('/home/sheridanmeth/Documents/markdown_examples/python/venv/bin/python') # Only use this if you don't want to use the default system python interpreter, such as if you don't have administrator privileges on your machine for installing packages. 
```


```{python, echo=FALSE, include=FALSE}
# Some WINDOWS users might need this if they are running python 
# using ANACONDA. And there are issues with the LaTeX interface when 
# using CONDA which are unresolved, such as 
# https://stackoverflow.com/q/58502671/5134817
# Run this after the r line library(reticulate)
#import os
#os.environ['QT_QPA_PLATFORM_PLUGIN_PATH'] = r'C:\Users\Harry\Anaconda3\Library\plugins\platforms'

# Some useful python modules
import matplotlib as mpl
import matplotlib.pyplot as plt
import pandas as pd
import numpy as np
import statsmodels.api as sm
import scipy as sp
from sklearn.linear_model import LinearRegression
from sklearn.feature_selection import f_regression
```

```{python, echo=FALSE, include=FALSE}
# Setting some default plotting features to give nicer plots. This can be commented out for beginners. 
rc_fonts = {
    "text.usetex": True,
    'text.latex.preview': True, # Gives correct legend alignment.
    'mathtext.default': 'regular',
    'figure.figsize': (6, 4),
    "font.family": "serif",
    "font.serif": "computer modern roman",
}
mpl.rcParams.update(rc_fonts)
```

# Abstract 

In this report we investigate the relationship between electricity prices
and daytime temperatures in Sydney Australia during 2013.
\todo{Perhaps we should update this data?}
We build a
second order non-linear model based on the minimum daytime temperature,
and predict the prices for the first week in January. 
\todo[inline=true]{Why do we not use more up to date data?}
This model is able to 
capture weekend and weekday differences. \todo{I must improve this report}

# The Raw Data

```{python, echo=FALSE, include=TRUE}
# Munging the Sydney temperature data. 
df = pd.read_csv('SydTemp.csv')
df = df[df['Year'] == 2013]
df.columns = [i.lower() for i in df.columns]
df['date'] = df[['year', 'month', 'day']].apply(lambda x: pd.datetime(*[int(i) for i in x]), axis=1)
df.columns = ['temp_max' if 'maximum' in i else 'temp_min' if 'minimum' in i else i for i in df.columns]
df = df[['date', 'temp_max', 'temp_min']]
# Munging the AEMO price data. 
df_prices = pd.read_csv("AEMOPriceDataSimple.csv")
df_prices.columns = [i.lower() for i in df_prices.columns]
df_prices['date'] = df_prices['date'].apply(pd.to_datetime)
# Merging the two datasets.
df = pd.merge(df, df_prices, on='date')
df = df.dropna()
df = df.sort_values(by='date')
# Making a dictionary for having some nice labels for printing to the markdown report or graphs.
column_names_to_labels = {'date':"Date", 'temp_max': "Maximum Temperature (C)", 'temp_min': "Minimum Temperature (C)", 'rrp': r'RRP (A\textdollar)'}  # This will be handy. 
# for WINDOWS users running ANACONDA cf. https://stackoverflow.com/q/58502671/5134817

# I produce a small data subset which I would like to print out. I save this into its own variable as I will need to copy this to an R dataframe object for producing a nice table using kable (sadly not possibly using python dataframes). 
df_example = df.head()
df_example = df_example.rename(columns=column_names_to_labels)
```



The data available is the RRP (the electricity price measured in AUD A\$), 
and the daily maximum and minimum temperature values, 
where some example data 
is shown in \Cref{tab:tab_example_data}.

```{r tab_example_data, echo=FALSE, include=TRUE}
example_data = py$df_example
kable(example_data, align='cccc', row.names = FALSE, caption = "\\label{tab:tab_example_data}Some example data illustrating the fields/features available for fitting.")
```


## Outliers

There is a notable outlier in the electricity prices for 20^th^ December 2013, 
(caused by an electrical storm). Due to the esoteric nature of such an event,
we do not want to consider such a point for use in everyday price predictions, 
so we exclude this from our considerations. The resultant dependence between the 
electricity prices and the minimum daily temperature is shown in 
 \Cref{fig:fig_prices_max_temp}. 

```{python, echo=FALSE, include=FALSE}
# we notice this has an obvious outlier in December
outliers = df[df['rrp'] > 200]
# we see this corresponds to the large electrical storm on 20th December 2013, so will delete
# this from our dataset.
df = df[df['rrp'] < 200]
```

```{python fig_prices_max_temp, echo=FALSE, include=TRUE, fig.cap="\\label{fig:fig_prices_max_temp}The dependence on the electricity prices with variations in the daily temperature minimum."}
plt.rcParams['figure.figsize'] = 7, 4  # If/usually you want to adjust the figure size from the default. 
plt.plot(df['temp_min'], df['rrp'], 'ko', markersize=2)
plt.xlabel(column_names_to_labels['temp_min'])
plt.ylabel(column_names_to_labels['rrp'])
plt.title("Electricity prices")
```

# Non-linear temperature effects 

We recall the linear model
\begin{equation}
\label{eqt:linear_model}
Y = \beta X + \epsilon.
\end{equation}

```{python, echo=FALSE, include=TRUE}
# For a multi variable linear regression using OLS
X = np.array([(df['temp_min'] ** i).values for i in [1,2]]).transpose()
X = sm.add_constant(X)
y = df['rrp'].values
linear_model_p_value = sm.OLS(y, X[:, :2]).fit().pvalues[-1] # The constant and linear term
non_linear_model_p_value = sm.OLS(y, X).fit().pvalues # All the features
```

```{r, echo=FALSE}
# We will want these floating point numbers (aka decimals using scientific notation) to be formatted when put inline with the text, so we can either format them here, or format when written inline (where we have opted for the latter).
linear_model_p_value <- py$linear_model_p_value
non_linear_model_p_value <- py$non_linear_model_p_value
```

To assess whether non-linear temperature effects are significant, we fit a 
linear model and a second order polynomial model to the data. We find the 
linear model \Cref{eqt:linear_model} has the $p$-value $`r signif(linear_model_p_value, 2)`$, 
 which is significant. Furthermore, trying a higher order non-linear model
with quadratic terms achieves $p$-values of $`r signif(non_linear_model_p_value, 2)`$, for the
intercept, linear, and quadratic terms respectively.
We can see these $p$-values are less than 0.05, and so we believe they are 
statistically significant to a 95\% confidence level (approximately a 
$2 \sigma$ 
confidence). 

\clearpage
# Building model matrices

We can build model matrices using the following R-function: 

```{python, echo=TRUE}
def pie_linear(x, k):
    """
    Description:
      Produces a model matrix for linear models using knots. 
    Input:
      x: Array, sample points.  
      k: Array, knot points.
    Return:
      Multidimensional array, model matrix. 
    """
    positive = lambda y: (y + np.fabs(y)) * 0.5
    model_matrix = np.zeros([len(x), 1 + len(k)])
    model_matrix[:, 0] = x
    for i in range(len(k)): 
        model_matrix[:, i + 1] = positive(x - k[i])
    return model_matrix
```

Which when given $\bm{x} = (1,2,\ldots,10)$ and knot points $(2,6)$ will output the
model matrix:

```{python, echo=TRUE}
x = np.linspace(1, 10, 10)
knots = np.array([2, 6])
X = pie_linear(x, knots)
print(X)
```

